---
title: "Reproducible Research: Peer Assessment 1"
author: Benjamin Hicks
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

The code for this analysis assumes that `activity.zip` is located in the local
git repo and then uses R's functions to extract it and load the csv to a data
frame

```{r}
knitr::opts_chunk$set(fig.path = "figure/")

setwd('~/RepData_PeerAssessment1/')
unzip('activity.zip')
fitness_df <- read.csv('activity.csv')
fitness_df <-na.omit(fitness_df) # omit NAs for now, unecessary because of how
                                 # aggregate functions but to be safe.
```

This produces a data frame with the following fields.

```{r}
head(fitness_df)
summary(fitness_df)
```


## What is mean total number of steps taken per day?

Producing a histogram requires getting all of the days in question by date 
and then creating sums of the total steps in each. I use the base R `aggregate`

```{r}
steps_day_df <- aggregate(steps~date, sum, data=fitness_df)
head(steps_day_df)

```

Some days had no data, so they are **not** included as imputed data points per 
instructions.

This produces a histogram as follows.
```{r}
hist(steps_day_df$steps, xlab='Steps Per Day', main='Histogram of Steps Taken Per Day (breaks of 5000)')
```

We see a peak between 10,000 - 15,000 steps per day. 

```{r}
mean_val <- round(mean(steps_day_df$steps))
med_val <- median(steps_day_df$steps)
```

This produced a mean of `r as.character(mean_val)` and median of `r med_val`.

## What is the average daily activity pattern?

To produce these values, we need to average all rows that have the same interval
and then aggregate.

```{r}
mean_steps_interval <- aggregate(steps~interval, mean, data=fitness_df)
summary(mean_steps_interval)

plot(mean_steps_interval, type='l', ylab='Steps', 
     xlab='Interval (in 5 minute increments)')
```

The time series plot clearly shows a peak approaching 1000 minutes. To confirm:
```{r}
mean_steps_interval[which(mean_steps_interval$steps == 
                          max(mean_steps_interval$steps)),]
```

The most steps occur on interval 835, which is around `r round(835/60)`:00 PM.

## Imputing missing values
The original data set has several rows with `NA`s for readings. This is probably 
skewing the statistics above. I therefore impute values for intervals based on
medians for a particular interval.

```{r}
fitness_df <- read.csv('activity.csv') # Now with NAs
sum(is.na(fitness_df$steps))
```

Using the data frame of means steps per interval recreated above, I impute any
NAs with the mean number of steps for that interval.

```{r}

imputed_df <- fitness_df
imputed_df$steps <- sapply(seq(1, nrow(fitness_df)), function (i) {
    if (is.na(fitness_df[i,1])) {
        mean_steps_interval[which(fitness_df[i,3] == mean_steps_interval$interval),2]
    }
    
    else {
        fitness_df[i,1]
    }
})

head(imputed_df)

```


I then repeat my procedure from the previous question on the imputed data.

```{r}
imputed_steps_day_df <- aggregate(steps~date, sum, data=imputed_df)
head(imputed_steps_day_df)

```

Producing the following histogram and averages:

```{r}
hist(imputed_steps_day_df$steps, xlab='Steps Per Day', main='Histogram of Steps Taken Per Day (breaks of 5000)')
mean_val <- round(mean(imputed_steps_day_df$steps))
med_val <- median(imputed_steps_day_df$steps)

mean_val
med_val
```
As we can see, there is little difference in the mean and median with the
imputed means, though the median goes up slightly.

## Are there differences in activity patterns between weekdays and weekends?

I first assign an extra column to set whether a day is a weekend or not.


```{r}
imputed_df$weekday <- sapply(seq(1, nrow(imputed_df)),
                                              function (i) {
    if (weekdays(as.Date(imputed_df[i,2])) %in% 
        c('Saturday', 'Sunday')) {
            as.factor('Weekend')
    }
                                                  
    else {
        as.factor('Weekday')
    }                                             
                                                  
})                                              

head(imputed_df)    
```
After this done, I aggregate using weekday along with interval and find means
for the two factors ('Weekday', 'Weekend').

```{r}
imputed_int_df <- aggregate(steps~interval+weekday, mean, data=imputed_df)
summary(imputed_int_df)
```

I then plot the graph into two time series panes using lattice.
```{r}
library(lattice)
xyplot(steps ~ interval | weekday,
       data = imputed_int_df,
       type = 'l',
       layout=c(1,2))
```

Based on the layout, we can see that on average, the data implies the test subject
moves much more on the same spike around interval 800 on weekdays.